```{r}
#| message: false
library(tidyverse)
library(scales)
library(kableExtra)

win_pred <- function(df){dbinom(1:df$matches,df$matches,df$prct_win)}
draw_pred <- function(df){dbinom(1:df$matches,df$matches,df$prct_draw)}

top_four <- tibble(club = c("Liverpool", "Aston_Villa", "Man_City", "Arsenal"),
                   matches_played = c(20,20,19,20),
                   matches = c(rep(18,2),19, 18),
                   wins = c(13,13,12,12),
                   draws = c(6,3,4,4),
                   loss = matches_played - (wins+draws),
                   pts = c(45,42,40,40),
                   prct_win = wins/matches_played,
                   prct_draw = draws/matches_played,
                   prct_loss = loss/matches_played)

t4_wins <- list()
t4_draws <- list()
t4_pred_wins <- NULL
t4_pred_draws <- NULL

for (i in 1:nrow(top_four)) {
  df <- top_four |> slice(i)
 t4_wins[[i]] <- win_pred(df)
 t4_draws[[i]] <- draw_pred(df)
 
 t4_pred_wins[i] <- which.max(t4_wins[[i]])
 t4_pred_draws[i] <- which.max(t4_draws[[i]])
}

preds_long <-  tibble(match = c(rep(1:18,2),1:19,1:18), 
                club = c(rep("Liverpool", 18), rep("Aston_Villa", 18), 
                         rep("Man_City", 19), rep("Arsenal", 18)),
                pred_wins = unlist(t4_wins),
                pred_draws = unlist(t4_draws))

plt_2024 <- preds_long |> 
  ggplot(aes(x = match, y = pred_wins, colour = club)) +
  geom_point(size = 3, shape = 20, position = position_jitter(width = 0.25), alpha = 0.3) + 
  scale_colour_manual(values = c("#EF0107","#670E36","#C8102E","lightblue")) +
  scale_y_continuous(labels = label_percent()) +
  scale_x_continuous(breaks = seq(1,19,1)) + 
  xlab("Matches") + 
  ylab("Chance") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  ggtitle("Predicted wins from remaining Premier League matches")

t4_tbl <- top_four |> mutate(pred_wins = t4_pred_wins,
             pred_draws = t4_pred_draws,
             pred_loss = matches-(pred_wins+pred_draws),
             pred_points = (3*pred_wins)+pred_draws,
             pred_total = pts + pred_points) |> select(-c(2,4:10))

matches <- 38
top_five <- tibble(club = c("Man_City", "Arsenal", "Man_Utd", "Newcastle", "Liverpool"),
                   matches = rep(38,5),
                   wins = c(28,26,23,19,19),
                   draws = c(5,6,6,14,10),
                   loss = matches - (wins + draws),
                   pts = c(89,84,75,71,67),
                   prct_win = wins/matches,
                   prct_draw = draws/matches,
                   prct_loss = loss/matches)

top_five_wins <- list()
top_five_draws <- list()
pred_wins <- NULL
pred_draws <- NULL

for (i in 1:nrow(top_five)) {
  df <- top_five |> slice(i)
 top_five_wins[[i]] <- win_pred(df)
 top_five_draws[[i]] <- draw_pred(df)
 
 pred_wins[i] <- which.max(top_five_wins[[i]])
 pred_draws[i] <- which.max(top_five_draws[[i]])
}



t5_tbl <- top_five |> select(Club = club, Matches = matches,
                   Wins = wins, Draws = draws, Losses = loss,
                   Points = pts, "% Won" = prct_win, "% Draw" = prct_draw,
                   "% Lost" = prct_loss) |> 
  kable(digits = 3) |> kable_styling("striped")

mc <- top_four |> filter(club == "Man_City") |> mutate(prct_win = 174/228, 
             prct_draw = 25/228)

liv <- top_four |> filter(club == "Liverpool") |> mutate(prct_win = 150/228, 
              prct_draw = 49/228)

mc_liv <- bind_rows(mc,liv)

mc_wins <- list()
mc_draws <- list()
pred_mc_wins <- NULL
pred_mc_draws <- NULL

for (i in 1:nrow(mc_liv)) {
  df <- mc_liv |> slice(i)
 mc_wins[[i]] <- win_pred(df)
 mc_draws[[i]] <- draw_pred(df)
 
 pred_mc_wins[i] <- which.max(mc_wins[[i]])
 pred_mc_draws[i] <- which.max(mc_draws[[i]])
}

mc_long <-  tibble(match = c(1:19,1:18),
                club = c(rep("Man_City", 19),rep("Liverpool", 18)),
                pred_wins = unlist(mc_wins),
                pred_draws = unlist(mc_draws))

# plt_mean <- mc_long |> 
#   ggplot(aes(x = match, y = pred_wins, colour = club)) +
#   geom_point(size = 3, shape = 20, position = position_jitter(width = 0.25), alpha = 0.5) + 
#   scale_colour_manual(values = c("#C8102E","lightblue")) +
#   scale_y_continuous(labels = label_percent()) +
#   scale_x_continuous(breaks = seq(1,19,1)) + 
#   xlab("Matches") + 
#   ylab("Chance") +
#   theme_minimal() +
#   theme(legend.title = element_blank()) +
#   ggtitle("Predicted wins from remaining Premier League matches")

tbl_mc_2024 <- mc_liv |> mutate(pred_wins = pred_mc_wins,
             pred_draws = pred_mc_draws,
             pred_loss = matches-(pred_wins+pred_draws),
             pred_points = (3*pred_wins)+pred_draws,
             pred_total = pts + pred_points) |> select(-c(2,4:10)) |> 
  arrange(desc(pred_total))
```

# Summary {#sec-summary}

At the half-way point in the 2023/24 Premier League season, I was reading the ESPN article [What's wrong at Arsenal?](https://www.espn.co.uk/football/insider/insider/story/_/id/39241757/explaining-arsenal-losses-investigation-stats-analysis). It's really good and full complicated analysis using expected goals, heatmaps and the inevitable nod to AI. It ends on Arsenal's chances of winning the league at 11% on January 5th 2024.

I'd previously read Oliver Johnson's Substack on [New Year's resolutions](https://bristoliver.substack.com/p/the-edge-is-everything) which included a binomial model of Ronny O'Sullivan's frame win and match win rate.

Just for fun, I wondered how a binomial model would compete with more sophisticated models for predicting the outcome of the second half of the season.

@sec-the-model has the details, but if the top four hold their form, then a binomial model predicts Liverpool to win the league as shown in @tbl-2024.

In the current season Manchester City with a 63% rate are performing well below their win rate of `r round(100*(174/228),0)`% and draw rate of `r round(100*(25/228),0)`% in the last six seasons.

Meanwhile Liverpool have a win rate of `r round(100*(150/228),0)`% and draw rate of `r round(100*(49/228),0)`% over the same period which is close their current 65% win rate.

Assuming there is regression to the mean, I also calculated @tbl-2024-mean for their six season mean win rates for Manchester City and Liverpool for the remaining matches. In this scenario Manchester City beat Liverpool by two points. 

Either way, it would be a surprise if Aston Villa won the league. And an even bigger one if Arsenal somehow won.

```{r}
#| label: tbl-2024
#| tbl-cap: Final predictions for 2023/24 season based on form from first half of the season.
t4_tbl |> select(Club = club, Matches = matches,
                   "Pred. Wins" = pred_wins, "Pred. Draws" = pred_draws, "Pred. Losses" = pred_loss,
                   "Pred. Points" = pred_points, "Pred. Total" = pred_total) |> 
  kable(digits = 3) |> kable_styling("striped")
```

```{r}
#| label: tbl-2024-mean
#| tbl-cap: Final predictions for 2023/24 season if Manchester City return to their long term form and the other clubs continue as the first half of the season.
tbl_mc_2024 |> select(Club = club, Matches = matches,
                   "Pred. Wins" = pred_wins, "Pred. Draws" = pred_draws, "Pred. Losses" = pred_loss,
                   "Pred. Points" = pred_points, "Pred. Total" = pred_total) |> 
  kable(digits = 3) |> kable_styling("striped")
```

## The model {#sec-the-model}

I got data for the 2022/23 and 2023/24 seasons from [Footystats](https://footystats.org/england/premier-league#)

```{r}
#| label: tbl-prem-league-2024
#| tbl-cap: Statistics for top four teams in 2023/24 season on January 10th 2024
top_four |> kable(digits = 3) |> kable_styling("striped")
```


```{r}
#| label: fig-2024
#| fig-cap: Binomial model of wins for remaining matches based on form from first half of the season
plt_2024
```
